Semi-Supervised Factored Logistic Regression for High-Dimensional Neuroimaging Data

### 2015-09-28 paper_nips/paper.tex
start of camera-ready

%\nipsfinalcopy % Uncomment for camera-ready version
\nipsfinalfalse
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

\begin{abstract}
Imaging neuroscience links human behavior to aspects of brain
biology in ever-increasing datasets.
%
Existing neuroimaging methods typically perform either discovery of unknown
neural structure or testing of neural structure associated with mental tasks.
%
However, testing hypotheses on the neural correlates of
larger mental task sets
necessitates adequate representations for the observations.
%
We therefore propose to blend
representation modelling and task classification into
a unified statistical learning problem.
%
A multinomial logistic regression is introduced that is
constrained by factored coefficients and coupled with an autoencoder.
%
We show that this approach yields more accurate and interpretable
neural models of psychological tasks in a reference dataset,
as well as better generalization to other datasets.
%

@@

In so doing, we search for the combined model parameters
$\theta=\{\mathbf{V_0,V_1,c_0, c_1, b_0, b_1}\}$
with respect to the (unsupervised) reconstruction error and the
(supervised) task classification.
${\mathcal{L_{AE}}}$ is devided by ${N_{X_{rest}}}$ to equilibrate both
loss terms to the same order of magnitude.
\(\Omega(\theta)\) represents an ElasticNet-type regularization
that combines $\ell_1$ and $\ell_2$ penalty terms.
%$\forall p \in \theta$.

@@

finding orthogonal network components using a change of basis
(whitening).

@@

initialized by random mixing matrix, whitening).

@@

(one hidden layer, $\ell_1=0.1$, $\ell_2=0.1$, 500 iterations).

@@

(two hidden layers, $\lambda=1$, $\ell_1=0.1$, $\ell_2=0.1$,

@@

\paragraph{Individual effects of dimensionality reduction and rest data.}
We first quantified the impact of introducing a bottleneck layer
disregarding the autoencoder.
To this end, ordinary logstic regression was juxtaposed
with SSFLogReg with $\lambda=1$.
For this experiment, we increased the difficulty of the classification problem
by including data from all 38 HCP tasks.
Indeed, increased class separability in component space,
as compared to voxel space, entails differences in
generalization performance of
$\approx{17\%}$ (Figure \ref{fig_dimred}).
%
Notably,
the cognitive tasks on reward and punishment processing
are among the least predicted with ordinary but well predicted with
SSFLogReg
(tasks 3 and 4 in Figure \ref{fig_dimred}).
These experimental conditions have been reported to exhibit
highly similar neural activity patterns in
GLM analyses of that dataset \cite{barch2013}.
Consequently, also local activity differences
(in the striatum and visual cortex in this case) can
be successfully captured by brain-network modelling.
%
\begin{figure}
\begin{centering}
\includegraphics[width=1.00\textwidth]{figures/accuracies.pdf}
\end{centering}
\vspace{-0.5cm}
\caption{\textbf{Effect of bottleneck in a 38-task classificaton problem}
Depicts the f1 score for each of 38 class predictions (chance is at $2,6\%$)
sorted from lowest to highest.
Multinomial logistic regression operating in voxel space (\textit{blue bars})
was compared to 
SSFLogReg operating in $20$
(\textit{left plot}) and $100$ (\textit{right plot})
latent modes (\textit{grey bars}).
Autoencoder or rest data were not used for these analyses
($\lambda=1$).
Ordinary logistic regression yielded 77,7\% accuracy out of sample, while
SSFLogReg scored at 94,4\% ($n=20$) and 94,2\% ($n=100$).
Hence, compressing the voxel data into a component space for classification
achieves higher task separability.
}
\label{fig_dimred}
\end{figure}

% rest effect
We then contemplated the impact of rest structure by modulating
its influence ($\lambda=0.25, 0.5, 0.75$) in data-scarce
and data-rich settings ($n=20$, $\ell_1=0.1$, $\ell_2=0.1$).
At the beginning of every epoch,
2000 task and 2000 rest maps were drawn with replacement
from same amounts of task and rest maps.
In frequently encountered data-scarce scenarios,
both training and validation
scores improved as we depart from the most supervised model ($lambda=1$).
This suggests that the autoencoder of SSFLogReg does not
behave like a classical regularization but more
like transfer learning.
In data-rich scenarios, was also present, although less pronounced.

but not in data-rich settings ($\approx{1000}$ labeled samples),
the repertoire of rest structure modulated model performance
(Figure \ref{fig_semisup}).

\begin{figure}[h]
\begin{centering}
\includegraphics[width=0.55\textwidth]{figures/semisup.png}
\vspace{-0.5cm}
\caption{\textbf{Effect of rest structure}
Model performance of SSFLogReg ($n=20$, $\ell_1=0.1$, $\ell_2=0.1$)
for different choices of $\lambda$ in
data-scarce (100 task and 100 rest maps, \textit{hot color}) and
data-rich (1000 task and 1000 task maps, \textit{cold color}) scenarios.
Gradient descent was performed on 2000 task and 2000 rest maps.
At the begining of each epoch, these were drawn with replacement
from a pool of 100 or 1000 different task and rest maps, respectively.
Especially the \textit{cold} curve
suggests a ``transfer learning''-type effect.
Chance is at 5,6\%.
}
\end{centering}
\label{fig_semisup}
\end{figure}

@@

\paragraph{Miscellaneous observations.}
For the sake of completeness,
we informally report modifications of the statistical model
that did not improve generalization performance.
%
$a)$
Introducing stochasticity into model learning by 
%$a)$ 
input corruption of $\mathbf{X_{task}}$ %or
%% $b)$ drop-out at the bottleneck $\mathbf{W_0}$ using binomial
%% distributions ($p=0.1, 0.3, 0.5$) 
deteriorated model performance
in all scenarios.
%
Adding $b)$ rectified linear units (ReLU) to $\mathbf{W_0}$ or
other commonly used nonlinearities ($c)$ sigmoid, $d)$ softplus,
$e)$ hyperbolic tangent) all led to decreased classification accuracies,
probably due to sample size limits.
%
Further, $f)$ ``pretraining'' of the bottleneck $\mathbf{W_0}$
(i.e., non-random initialization) by 
either corresponding PCA, SPCA or ICA loadings did not exhibit improved accuracies,
neither did $h)$ autoencoder pretraining.
%
Moreover,
introducing an additional $h)$ overcomplete layer (100 units)
after the bottleneck was not advantageous.
%
Finally, imposing either $i)$ only $\ell_1$ or 
$k)$ only $\ell_2$ penalty terms
was disadvantageous in all tested cases.
Thiss favored ElasticNet regularization chosen in the above analyses.

>>>>

% \nipsfinalcopy % Uncomment for camera-ready version
\nipsfinaltrue
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

\begin{abstract}
Imaging neuroscience links human behavior to aspects of brain
biology in ever-increasing datasets.
%
Existing neuroimaging methods typically perform either discovery of unknown
neural structure or testing of neural structure associated with mental tasks.
%
However, testing hypotheses on the neural correlates underlying
larger sets of mental tasks
necessitates adequate representations for the observations.
%
We therefore propose to blend
representation modelling and task classification into
a unified statistical learning problem.
%
A multinomial logistic regression is introduced that is
constrained by factored coefficients and coupled with an autoencoder.
%
We show that this approach yields more accurate and interpretable
neural models of psychological tasks in a reference dataset,
as well as better generalization to other datasets.
%

@@

In so doing, we search for the combined model parameters
$\theta=\{\mathbf{V_0,V_1,c_0, c_1, b_0, b_1}\}$
with respect to the (unsupervised) reconstruction error and the
(supervised) task detection.
${\mathcal{L_{AE}}}$ is devided by ${N_{X_{rest}}}$ to equilibrate both
loss terms to the same order of magnitude.
\(\Omega(\theta)\) represents an ElasticNet-type regularization
that combines $\ell_1$ and $\ell_2$ penalty terms.
%$\forall p \in \theta$.

@@

finding orthogonal network components (whitening of the data).

@@

initialized by random mixing matrix, whitening of the data).

@@

(no hidden layer, $\ell_1=0.1$, $\ell_2=0.1$, 500 iterations).

@@

(one hidden layers, $\lambda=1$, $\ell_1=0.1$, $\ell_2=0.1$,

@@

\paragraph{Individual effects of dimensionality reduction and rest data.}
We first quantified the impact of introducing a bottleneck layer
disregarding the autoencoder.
To this end, ordinary logistic regression was juxtaposed
with SSFLogReg with $\lambda=1$.
For this experiment, we increased the difficulty of the classification problem
by including data from all 38 HCP tasks.
Indeed, increased class separability in component space,
as compared to voxel space, entails differences in
generalization performance of
$\approx{17\%}$ (Figure \ref{fig_dimred}).
%
Notably,
the cognitive tasks on reward and punishment processing
are among the least predicted with ordinary but well predicted with
SSFLogReg
(tasks 3 and 4 in Figure \ref{fig_dimred}).
These experimental conditions have been reported to exhibit
highly similar neural activity patterns in
GLM analyses of that dataset \cite{barch2013}.
Consequently, also local activity differences
(in the striatum and visual cortex in this case) can
be successfully captured by brain-network modelling.
%
\begin{figure}
\begin{centering}
\includegraphics[width=1.00\textwidth]{figures/accuracies.pdf}
\end{centering}
\vspace{-0.5cm}
\caption{\textbf{Effect of bottleneck in a 38-task classificaton problem}
Depicts the f1 score for each of 38 class predictions (chance is at $2,6\%$)
sorted from lowest to highest.
Multinomial logistic regression operating in voxel space (\textit{blue bars})
was compared to 
SSFLogReg operating in $20$
(\textit{left plot}) and $100$ (\textit{right plot})
latent modes (\textit{grey bars}).
Autoencoder or rest data were not used for these analyses
($\lambda=1$).
Ordinary logistic regression yielded 77,7\% accuracy out of sample, while
SSFLogReg scored at 94,4\% ($n=20$) and 94,2\% ($n=100$).
Hence, compressing the voxel data into a component space for classification
achieves higher task separability.
}
\label{fig_dimred}
\end{figure}

% rest effect
We then contemplated the impact of rest structure by modulating
its influence ($\lambda=0.25, 0.5, 0.75$) in data-scarce
and data-rich settings ($n=20$, $\ell_1=0.1$, $\ell_2=0.1$).
At the beginning of every epoch,
2000 task and 2000 rest maps were drawn with replacement
from same amounts of task and rest maps.
In data-scarce scenarios (frequently encountered by neuroimaging practitioners),
the out-of-sample scores improve
as we depart from the most supervised model ($lambda=1$).
This suggests that the autoencoder of SSFLogReg does not
behave like a classical regularization but more
like transfer learning.
In data-rich scenarios, we observed the same trend to be apparent.

\begin{figure}[h]
\begin{centering}
\includegraphics[width=0.55\textwidth]{figures/semisup.png}
\vspace{-0.5cm}
\caption{\textbf{Effect of rest structure}
Model performance of SSFLogReg ($n=20$, $\ell_1=0.1$, $\ell_2=0.1$)
for different choices of $\lambda$ in
data-scarce (100 task and 100 rest maps, \textit{hot color}) and
data-rich (1000 task and 1000 task maps, \textit{cold color}) scenarios.
Gradient descent was performed on 2000 task and 2000 rest maps.
At the begining of each epoch, these were drawn with replacement
from a pool of 100 or 1000 different task and rest maps, respectively.
Especially the \textit{hot} curve
suggests a ``transfer learning''-type effect.
Chance is at 5,6\%.
}
\end{centering}
\label{fig_semisup}
\end{figure}

@@

\paragraph{Miscellaneous observations.}
For the sake of completeness,
we informally report modifications of the statistical model
that did not improve generalization performance.
%
$a)$
Introducing stochasticity into model learning by 
%$a)$ 
input corruption of $\mathbf{X_{task}}$ %or
%% $b)$ drop-out at the bottleneck $\mathbf{W_0}$ using binomial
%% distributions ($p=0.1, 0.3, 0.5$) 
deteriorated model performance
in all scenarios.
%
Adding $b)$ rectified linear units (ReLU) to $\mathbf{W_0}$ or
other commonly used nonlinearities ($c)$ sigmoid, $d)$ softplus,
$e)$ hyperbolic tangent) all led to decreased classification accuracies,
probably due to sample size limits.
%
Further, $f)$ ``pretraining'' of the bottleneck $\mathbf{W_0}$
(i.e., non-random initialization) by 
either corresponding PCA, SPCA or ICA loadings did not exhibit improved accuracies,
neither did $h)$ autoencoder pretraining.
%
Moreover,
introducing an additional $h)$ overcomplete layer (100 units)
after the bottleneck was not advantageous.
%
Finally, imposing either $i)$ only $\ell_1$ or 
$k)$ only $\ell_2$ penalty terms
was disadvantageous in all tested cases.
This favored ElasticNet regularization chosen in the above analyses.
